Do investors in clean energy ETFs herd? The role of climate risks

### 2025-06-29 Journal/journal_submission/alt_energy_herding_no_author.tex
second journal submission

This study offers new insights into the herding behaviour in US clean energy Exchange-traded funds (ETFs) over the period from May 1, 2016 to June 19, 2024. 
The baseline model shows significant herding. An extended mode indicates that herding is present in both down and up markets, with a stronger effect in the down market, suggesting an asymmetry. 
Herding is also found to be time-varying. Further analysis shows that the transition climate risk, particularly its high level, reduces the probability of herding in clean energy ETFs, whereas physical climate risk does not exert any significant impact on the probability of herding. 
Thus, large levels of transition climate risk seem to encourage market efficiency in clean energy ETFs  and climate-hedging behavior by investors.

@@

While herding behaviour has been examined in ESG markets, it remains
understudied in clean energy assets, notably, clean energy
ETFs.\footnote{Amongst others, \citet{loang2023} shows that compliance
  with SDG goals can introduce bias in investor sentiment, which leads
  to herding behaviour. Using a Twitter (or X) uncertainty index,
  \citet{koutmos2024} finds evidence of herding in US-based ESG index
  fund investors. \citet{przychodzen2016} found indicate evidence of
  herding behaviour amongst fund managers who incorporated ESG
  strategies in their portfolios. Lastly, \citet{rubbaniy2021} highlight
  evidence of herding in the MSCI US ESG Leader Index during extreme
  (bear and bull periods) periods.} Other studies on herding have been
conducted in commodity and fossil energy markets. For example,
\citet{demirer2013} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} shows herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} find significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} show that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} find no herding behaviour in
international commodity markets. Notably, our study extends the results
of \citet{dragomirescu-gaina2021} who have examined herding behavior of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They study the energy
equities included in the S\&P 500 and concluded that herding among
investors in the US energy market sector is sensitive to green
volatility shocks

@@

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behavior can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep[e.g.][]{lakonishok1992}, and studies that use market returns data
and investigate herding towards the market consensus
\citep[e.g.][]{chang2000, galariotis2015}. Our paper falls within the
latter category and tests for herding towards the market consensus for
clean energy US ETFs.

@@

It is expected that environmentally conscious investors would prefer to
hold clean energy assets that perform well in the face of increasing
climate change risks \citep[see][]{bouri2022}, even if this entails
accepting lower returns for such climate-hedging assets. Therefore, in
the context of our study and following previous studies on the
determinants of herding behaviour \citep[see][]{bouri2019, demirer2018},
we examine the effect of physical and transition climate risks on the
formation of herding behavior in the clean energy ETF market.

Furthermore, higher transition risk means greater uncertainty for future
pathway towards a cleaner economy. Under these conditions, green assets
such as green stocks or ETFs are expected to perform better and can
serve as hedging tools \citep[see inter alia][]{pastor}. As a result,
investors allocate their capital to available green assets thus
resulting in higher returns' dispersion with respect to the aggregate
market. Stated differently, investors move their capital away from
market consensus across green ETFs ruling out the possibility of herding
behavior.

>>>>

This study offers new insights into the herding behaviour in US clean energy Exchange-traded funds (ETFs) over the period from May 1, 2016 to June 19, 2024. 
The baseline model shows significant herding. An extended mode indicates that herding is present in both down and up markets, with a stronger effect in the down market, suggesting an asymmetry. 
Herding is also found to be time-varying. Further analysis shows that the transition climate risk, particularly its high level, reduces the probability of herding in clean energy ETFs, whereas physical climate risk does not exert any significant impact on the probability of herding. 
Thus, large levels of transition climate risk seem to encourage market efficiency in clean energy ETFs  and climate-hedging behaviour by investors.

@@

While herding behaviour has been examined in ESG markets, it remains
understudied in clean energy assets, notably, clean energy
ETFs.\footnote{Amongst others, \citet{loang2023} shows that compliance
  with SDG goals can introduce bias in investor sentiment, which leads
  to herding behaviour. Using a Twitter (or X) uncertainty index,
  \citet{koutmos2024} finds evidence of herding in US-based ESG index
  fund investors. \citet{przychodzen2016} found indicate evidence of
  herding behaviour amongst fund managers who incorporated ESG
  strategies in their portfolios. Lastly, \citet{rubbaniy2021} highlight
  evidence of herding in the MSCI US ESG Leader Index during extreme
  (bear and bull periods) periods.} Other studies on herding have been
conducted in commodity and fossil energy markets. For example,
\citet{demirer2013} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} shows herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} find significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} show that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} find no herding behaviour in
international commodity markets. Notably, our study extends the results
of \citet{dragomirescu-gaina2021} who have examined herding behaviour of
investors in the US energy sector and herding sensitivity to various
proxies of policy uncertainty and financial risk. They study the energy
equities included in the S\&P 500 and concluded that herding among
investors in the US energy market sector is sensitive to green
volatility shocks

@@

It is well established that herding literature is vast with
contradictory results depending mainly on the market, the employed
methodology and the period under consideration \citep{spyrou2013}.
Herding behaviour can be either spurious in cases when investors make
similar decisions as a result of processing the same information set and
intentional herding when investors imitate the actions of others
\citep[see inter alia][]{bikhchandani2000, galariotis2015}. Empirical
studies on herding usually fall into two categories: namely those that
employ holdings data aiming at measuring institutional investor herding
\citep[e.g.][]{lakonishok1992}, and studies that use market returns data
and investigate herding towards the market consensus
\citep[e.g.][]{chang2000, galariotis2015}. Our paper falls within the
latter category and tests for herding towards the market consensus for
clean energy US ETFs.

@@

It is expected that environmentally conscious investors would prefer to
hold clean energy assets that perform well in the face of increasing
climate change risks \citep[see][]{bouri2022}, even if this entails
accepting lower returns for such climate-hedging assets. Therefore, in
the context of our study and following previous studies on the
determinants of herding behaviour \citep[see][]{bouri2019, demirer2018},
we examine the effect of physical and transition climate risks on the
formation of herding behaviour in the clean energy ETF market.

Furthermore, higher transition risk means greater uncertainty for future
pathway towards a cleaner economy. Under these conditions, green assets
such as green stocks or ETFs are expected to perform better and can
serve as hedging tools \citep[see inter alia][]{pastor}. As a result,
investors allocate their capital to available green assets thus
resulting in higher returns' dispersion with respect to the aggregate
market. Stated differently, investors move their capital away from
market consensus across green ETFs ruling out the possibility of herding
behaviour.

### 2025-06-29 alt_energy_herding.qmd
second journal submission

While herding behaviour has been examined in ESG markets, it remains understudied in clean energy assets, notably, clean energy ETFs.^[Amongst others, @loang2023 shows that compliance with SDG goals can introduce bias in investor sentiment, which leads to herding behaviour. Using a Twitter (or X) uncertainty index, @koutmos2024 finds evidence of herding in US-based ESG index fund investors. @przychodzen2016 found indicate evidence of herding behaviour amongst fund managers who incorporated ESG strategies in their portfolios. Lastly, @rubbaniy2021 highlight evidence of herding in the MSCI US ESG Leader Index during extreme (bear and bull periods) periods.] Other studies on herding have been conducted in commodity and fossil energy markets. For example, @demirer2013 conducted a commodity sectoral study and found herding behaviour in grains but not in other sectors. Similarly, @gilbert2010 shows herding behaviour amongst speculators in non-ferrous commodities. Others did not find evidence of herding in similar markets. @babalos2015 find significant anti-herding behaviour in metal commodities futures after the global financial crisis. @pierdzioch2010 show that forecasters in oil and metals markets deviated from the crowd, indicating a rational response to market information. @steen2013 find no herding behaviour in international commodity markets. Notably,  our study extends the results of @dragomirescu-gaina2021 who have examined herding behavior of investors in the US energy sector and herding sensitivity to various proxies of policy uncertainty and financial risk. They study the energy equities included in the S&P 500 and concluded that herding among investors in the US energy market sector is sensitive to green volatility shocks

@@

It is well established that herding literature is vast with contradictory results depending mainly on the market, the employed methodology and the period under consideration [@spyrou2013]. Herding behavior can be either spurious in cases when investors make similar decisions as a result of processing the same information set and intentional herding when investors imitate the actions of others [see inter alia @bikhchandani2000; @galariotis2015]. Empirical studies on herding usually fall into two categories: namely those that employ holdings data aiming at measuring institutional investor herding [e.g. @lakonishok1992], and studies that use market returns data and investigate herding towards the market consensus [e.g. @chang2000; @galariotis2015]. Our paper falls within the latter category and tests for herding towards the market consensus for clean energy US ETFs.

@@

It is expected that environmentally conscious investors would prefer to hold clean energy assets that perform well in the face of increasing climate change risks [see @bouri2022], even if this entails accepting lower returns for such climate-hedging assets. Therefore, in the context of our study and following previous studies on the determinants of herding behaviour [see @bouri2019; @demirer2018], we examine the effect of physical and transition climate risks on the formation of herding behavior in the clean energy ETF market.

Furthermore, higher transition risk means greater uncertainty for future pathway towards a cleaner economy. Under these conditions, green assets such as green stocks or ETFs are expected to perform better and can serve as hedging tools [see inter alia @pastor]. As a result, investors allocate their capital to available green assets thus resulting in higher returns’ dispersion with respect to the aggregate market. Stated differently, investors move their capital away from market consensus across green ETFs ruling out the possibility of herding behavior.

>>>>

While herding behaviour has been examined in ESG markets, it remains understudied in clean energy assets, notably, clean energy ETFs.^[Amongst others, @loang2023 shows that compliance with SDG goals can introduce bias in investor sentiment, which leads to herding behaviour. Using a Twitter (or X) uncertainty index, @koutmos2024 finds evidence of herding in US-based ESG index fund investors. @przychodzen2016 found indicate evidence of herding behaviour amongst fund managers who incorporated ESG strategies in their portfolios. Lastly, @rubbaniy2021 highlight evidence of herding in the MSCI US ESG Leader Index during extreme (bear and bull periods) periods.] Other studies on herding have been conducted in commodity and fossil energy markets. For example, @demirer2013 conducted a commodity sectoral study and found herding behaviour in grains but not in other sectors. Similarly, @gilbert2010 shows herding behaviour amongst speculators in non-ferrous commodities. Others did not find evidence of herding in similar markets. @babalos2015 find significant anti-herding behaviour in metal commodities futures after the global financial crisis. @pierdzioch2010 show that forecasters in oil and metals markets deviated from the crowd, indicating a rational response to market information. @steen2013 find no herding behaviour in international commodity markets. Notably,  our study extends the results of @dragomirescu-gaina2021 who have examined herding behaviour of investors in the US energy sector and herding sensitivity to various proxies of policy uncertainty and financial risk. They study the energy equities included in the S&P 500 and concluded that herding among investors in the US energy market sector is sensitive to green volatility shocks

@@

It is well established that herding literature is vast with contradictory results depending mainly on the market, the employed methodology and the period under consideration [@spyrou2013]. Herding behaviour can be either spurious in cases when investors make similar decisions as a result of processing the same information set and intentional herding when investors imitate the actions of others [see inter alia @bikhchandani2000; @galariotis2015]. Empirical studies on herding usually fall into two categories: namely those that employ holdings data aiming at measuring institutional investor herding [e.g. @lakonishok1992], and studies that use market returns data and investigate herding towards the market consensus [e.g. @chang2000; @galariotis2015]. Our paper falls within the latter category and tests for herding towards the market consensus for clean energy US ETFs.

@@

It is expected that environmentally conscious investors would prefer to hold clean energy assets that perform well in the face of increasing climate change risks [see @bouri2022], even if this entails accepting lower returns for such climate-hedging assets. Therefore, in the context of our study and following previous studies on the determinants of herding behaviour [see @bouri2019; @demirer2018], we examine the effect of physical and transition climate risks on the formation of herding behaviour in the clean energy ETF market.

Furthermore, higher transition risk means greater uncertainty for future pathway towards a cleaner economy. Under these conditions, green assets such as green stocks or ETFs are expected to perform better and can serve as hedging tools [see inter alia @pastor]. As a result, investors allocate their capital to available green assets thus resulting in higher returns’ dispersion with respect to the aggregate market. Stated differently, investors move their capital away from market consensus across green ETFs ruling out the possibility of herding behaviour.

### 2025-06-29 alt_energy_herding.tex
second journal submission

Green energy sectors like solar, wind, geothermal, biofuels, biomass,
wave, hydro and tidal energies are considered by investors, authorities
and policy-makers as the substitutes for fossil-fuels energy that
increase CO2 emissions \citep{shahzad2020clean}. Moreover, as
\citet{naqvi2022} point out that ``the alternative energy sector is
expected to grow faster over the next decades due to the depletion of
fossil fuel reserves and the technological advance that has made green
energy feasible''.

>>>>

Green energy sectors like solar, wind, geothermal, biofuels, biomass,
wave, hydro and tidal energies are considered by investors, authorities
and policy-makers as the substitutes for fossil-fuels energy that
increase \(CO_2\) emissions \citep{shahzad2020clean}. Moreover, as
\citet{naqvi2022} point out that ``the alternative energy sector is
expected to grow faster over the next decades due to the depletion of
fossil fuel reserves and the technological advance that has made green
energy feasible''.
